---
 title: "RepeatedCircadian tutorial"
 auther: "Zhiguang Huo, Haocheng Ding (Department of Biostatistics, University of Florida)"
 date: "`r Sys.Date()`"
 output: 
   html_document:
    toc: true
    toc_depth: 2
    number_sections: true
 vignette: >
  %\VignetteIndexEntry{RepeatedCircadian}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(stringsAsFactors = FALSE)
```

# Circadian Rhythmicity Detection

## Read in the example data
```{r}
library(lme4)
library(pbkrtest)
library(lmtest)
library(RepeatedCircadian) # include RepeatedCircadian package
# load the data
data("HTRF")
data("pheno")
```

## Performing Circadian Rhythmicity Detection for all genes in BA11 data

```{r}
ngene <- nrow(HTRF)
sample <- colnames(HTRF)
sample_split <- strsplit(sample, split=".", fixed=T)
ID <- unlist(lapply(sample_split, function(x) x[1]))
condition <- unlist(lapply(sample_split, function(x) x[2]))
time <- unlist(lapply(sample_split, function(x) x[3]))
sample_df <- data.frame(ID, condition, time)
sample_df$Time <- 1
sample_df$Time[which(sample_df$time=="T5")] <- 5
sample_df$Time[which(sample_df$time=="T9")] <- 9
sample_df$Time[which(sample_df$time=="T13")] <- 13
sample_df$Time[which(sample_df$time=="T17")] <- 17
sample_df$Time[which(sample_df$time=="T19")] <- 19

indR <- which(sample_df$condition=="R")  # 63
indU <- which(sample_df$condition=="U")  # 62




# Real Data circadian pattern detection 
ttR <- sample_df$Time[indR] # Circadian time of restricted group
ttU <- sample_df$Time[indU] # Circadian time of unrestricted group
idR <- sample_df$ID[indR] # ID of restricted group
idU <- sample_df$ID[indU] # ID of unrestricted group


tableR <- matrix(nrow=ngene, ncol=7)
for (i in 1:ngene){
  yy <- as.numeric(HTRF[i, indR])
  aTest <- rpt_rhythmicity(ttR, yy, idR, period=24, method="LR")
  tableR[i,] <- unlist(aTest)
}
Symbol <- row.names(HTRF)
tableR <- data.frame(Symbol, tableR)
colnames(tableR)<-c("Symbol","statistic","pvalue","amplitude","phi","basal","sigma_alpha","sigma_0")
p.value <- tableR$pvalue
q.value <- p.adjust(p.value, method = "BH", n = length(p.value))
tableR <- data.frame(tableR, qvalue = q.value)
colnames(tableR)<-c("Symbol","statistic","pvalue","amplitude","phi","basal","sigma_alpha","sigma_0","qvalue")
tableR$r <- tableR$amplitude/tableR$sigma_alpha # (sigmaA2: residual variance of the circadian model)


tableU <- matrix(nrow=ngene, ncol=7)
for (i in 1:ngene){
  yy <- as.numeric(HTRF[i, indU])
  aTest <- rpt_rhythmicity(ttU, yy, idU, period=24, method="LR")
  tableU[i,] <- unlist(aTest)
}
Symbol <- row.names(HTRF)
tableU <- data.frame(Symbol, tableU)

colnames(tableU)<-c("Symbol","statistic","pvalue","amplitude","phi","basal","sigma_alpha","sigma_0")
p.value <- tableU$pvalue
q.value <- p.adjust(p.value, method = "BH", n = length(p.value))
tableU <- data.frame(tableU, qvalue = q.value)
colnames(tableU)<-c("Symbol","statistic","pvalue","amplitude","phi","basal","sigma_alpha","sigma_0","qvalue")
tableU$r <- tableU$amplitude/tableU$sigma_alpha # (sigmaA2: residual variance of the circadian model)
```

## Save the Circadian Rhythmicity result as csv files.
```{r, eval = FALSE}
write.csv(tableR, "Table_pattern_Restricted_notorder.csv", row.names = F)
write.csv(tableU, "Table_pattern_Unrestricted_notorder.csv", row.names = F)
oR <- order(tableR$pvalue)
otableR <- tableR[oR,]
oU <- order(tableU$pvalue)
otableU <- tableU[oU,]
write.csv(otableR, "Table_pattern_Restricted.csv", row.names = F)
write.csv(otableU, "Table_pattern_Unrestricted.csv", row.names = F)
```


## Plot top significant genes with circadian rhythmicity
```{r}

## for visualization purpose
sinCurve <- function(x, A, phase, basal){
	A24 <- 24
	A * sin(2*pi/A24 * (x + phase)) + basal
}

tt_base <- seq(-6,24,0.1)

# Restricted group top 4 significant genes plots
topN <- 4
par(mfrow = c(2, 2))
geneName <- c("ARNTL","CIART","HPB1","KLHL25")
for(i in 1:topN){
	this_LR <- otableR[i,]
	yy_obs <- HTRF[this_LR$Symbol, indR]
	rptout <- rpt_rhythmicity(ttR,as.numeric(HTRF[this_LR$Symbol, indR]),idR)
	yy_fit <- sinCurve(tt_base, rptout$A, rptout$phi, rptout$basal)
	#yy_fit <- sinCurve(tt_base, this_LR$amplitude, this_LR$phi, this_LR$basal)
	atitle <- paste0(geneName[i], ", p = ", signif(this_LR$pvalue,3))	
	plot(ttR, yy_obs ,pch=20, xlab = "Circadian Time", ylab="Expression Value", main = atitle)
	lines(tt_base, yy_fit, col = 2, lwd =3)
}
```

# Differential Circadian Pattern Detection

## Data preparation for differential circadian analysis
We first create two datasets based on two age groups: young ($\le 40$ years) group and old ($\ge 60$ years) group.
We will then perform differential circadian analysis comparing young and old groups.

```{r}
pheno_old <- pheno[pheno$age >= 60, ]
BA11_old <- BA11[,pheno_old$ID]
tt_old <- pheno_old$tod

pheno_young <- pheno[pheno$age <= 40, ]
BA11_young <- BA11[,pheno_young$ID]
tt_young <- pheno_young$tod
```

We also performed circadian rhythmicity detection in young and old groups, respectively, to facilitate further differential circadian analysis.

```{r}
## circadian analysis in the old group
thisResult_old <- data.frame(genes = rownames(BA11_old), 
	amp =NA, phase = NA, peakTime = NA, basal = NA, pvalue = NA
)

gg <- 1
for(gg in 1:nrow(thisResult_old)){
	if(gg %% 1000 == 0){
		cat('processed', gg, 'genes\n')
	}
	yy_old <- as.numeric(BA11_old[gg, ]) ## make sure yy is a numeric vector instead of dataframe. 
	aLR <- LR_rhythmicity(tt_old, yy_old)
	
	thisResult_old$amp[gg] <- aLR$amp
	thisResult_old$phase[gg] <- aLR$phase
	thisResult_old$peakTime[gg] <- aLR$peakTime
	thisResult_old$basal[gg] <- aLR$offset
	thisResult_old$pvalue[gg] <- aLR$pvalue
}


## circadian analysis in the young group
thisResult_young <- data.frame(genes = rownames(BA11_young), 
	amp =NA, phase = NA, peakTime = NA, basal = NA, pvalue = NA
)

gg <- 1
for(gg in 1:nrow(thisResult_young)){
	if(gg %% 1000 == 0){
		cat('processed', gg, 'genes\n')
	}
	yy_young <- as.numeric(BA11_young[gg, ]) ## make sure yy is a numeric vector instead of dataframe. 
	aLR <- LR_rhythmicity(tt_young, yy_young)
	
	thisResult_young$amp[gg] <- aLR$amp
	thisResult_young$phase[gg] <- aLR$phase
	thisResult_young$peakTime[gg] <- aLR$peakTime
	thisResult_young$basal[gg] <- aLR$offset
	thisResult_young$pvalue[gg] <- aLR$pvalue
}

```


## Differential amplitude, differential phase, and differential basal level

### Candidate gene selection
The prerequisite for differential amplitude, differential phase, and  differential  basal  level  is  that  there  should  exist circadian  rhythmicity  in  both  young and old groups under comparisons.
Therefore, we suggested users to set $p \le 0.05$ from our previous likelihood-based test to ensure the existence of the circadian rhythmicity in both conditions.

```{r}
## genes that have circadian rhythmicity in both young and old groups
logic_both <- thisResult_old$pvalue<0.05 & thisResult_young$pvalue<0.05
genes_both <- thisResult_old$genes[logic_both]

BA11_old23 <- BA11_old[genes_both,]
BA11_young23 <- BA11_young[genes_both,]
```


### Differential amplitude

**Perform differential amplitude**
```{r}
result_diff_amp <- data.frame(genes = genes_both,
		amp_young = NA,
		amp_old = NA, 
		amp_diff = NA, 
		pvalue = NA
		)

for(gg in 1:nrow(result_diff_amp)){
  ayy_young <- BA11_young23[gg, ]
  ayy_old <- BA11_old23[gg, ]

  adiff_amp <- LR_diff(tt_young,ayy_young,tt_old,ayy_old,type="amplitude") # 

  result_diff_amp$amp_young[gg] <- adiff_amp$amp_1
  result_diff_amp$amp_old[gg] <- adiff_amp$amp_2
  result_diff_amp$amp_diff[gg] <- adiff_amp$amp_2 - adiff_amp$amp_1
  result_diff_amp$pvalue[gg] <- adiff_amp$pvalue
}

result_diff_amp_sorted <- result_diff_amp[order(result_diff_amp$pvalue), ]
```

**Save the result**

```{r, eval = FALSE}
write.csv(result_diff_amp_sorted, file = "result_diff_amp_sorted.csv", row.names = FALSE)
```

**Visualize the most significant gene**

```{r fig1, fig.height = 4, fig.width = 10}
topGene_diffamp <- result_diff_amp_sorted$genes[1]
byy_young <- BA11_young23[topGene_diffamp, ]
aLR_young <- LR_rhythmicity(tt_young, byy_young)
byy_young_fit <- sinCurve(tt_base, aLR_young$amp, aLR_young$phase, aLR_young$offset)
	
byy_old <- BA11_old23[topGene_diffamp, ]
aLR_old <- LR_rhythmicity(tt_old, byy_old)
byy_old_fit <- sinCurve(tt_base, aLR_old$amp, aLR_old$phase, aLR_old$offset)

ymax <- max(c(byy_young, byy_old))
ymin <- min(c(byy_young, byy_old))

title_young <- paste0("young group, amp = ", signif(result_diff_amp_sorted$amp_young[1],2))
title_old <- paste0("old group, amp = ", signif(result_diff_amp_sorted$amp_old[1],2))

par(mfrow = c(1,2))
plot(tt_young, byy_young ,pch=20, xlab = "Time of Death (TOD)", ylab="Expression Value", main = title_young, ylim = c(ymin, ymax))
lines(tt_base, byy_young_fit, col = 2, lwd =3)

plot(tt_old, byy_old ,pch=20, xlab = "Time of Death (TOD)", ylab="Expression Value", main = title_old)
lines(tt_base, byy_old_fit, col = 4, lwd =3)
```


### Differential phase

**Perform differential phase analysis**
```{r}
result_diff_phase <- data.frame(genes = genes_both,
		phase_young = NA,
		phase_old = NA, 
		phase_diff = NA, 
		pvalue = NA
		)

for(gg in 1:nrow(result_diff_phase)){
  ayy_young <- BA11_young23[gg, ]
  ayy_old <- BA11_old23[gg, ]

  adiff_phase <- LR_diff(tt_young,ayy_young,tt_old,ayy_old,type="phase") # 

  result_diff_phase$phase_young[gg] <- adiff_phase$phase_1
  result_diff_phase$phase_old[gg] <- adiff_phase$phase_2
  result_diff_phase$phase_diff[gg] <- adiff_phase$phase_2 - adiff_phase$phase_1
  result_diff_phase$pvalue[gg] <- adiff_phase$pvalue
}

result_diff_phase_sorted <- result_diff_phase[order(result_diff_phase$pvalue), ]
```

**Save the result**

```{r, eval = FALSE}
write.csv(result_diff_phase_sorted, file = "result_diff_phase_sorted.csv", row.names = FALSE)
```

**Visualize the most significant gene**

```{r fig2, fig.height = 4, fig.width = 10}
topGene_diffphase <- result_diff_phase_sorted$genes[1]
byy_young <- BA11_young23[topGene_diffphase, ]
aLR_young <- LR_rhythmicity(tt_young, byy_young)
byy_young_fit <- sinCurve(tt_base, aLR_young$amp, aLR_young$phase, aLR_young$offset)
	
byy_old <- BA11_old23[topGene_diffphase, ]
aLR_old <- LR_rhythmicity(tt_old, byy_old)
byy_old_fit <- sinCurve(tt_base, aLR_old$amp, aLR_old$phase, aLR_old$offset)

ymax <- max(c(byy_young, byy_old))
ymin <- min(c(byy_young, byy_old))

title_young <- paste0("young group, phase = ", signif(result_diff_phase_sorted$phase_young[1],2))
title_old <- paste0("old group, phase = ", signif(result_diff_phase_sorted$phase_old[1],2))

par(mfrow = c(1,2))
plot(tt_young, byy_young ,pch=20, xlab = "Time of Death (TOD)", ylab="Expression Value", main = title_young, ylim = c(ymin, ymax))
lines(tt_base, byy_young_fit, col = 2, lwd =3)

plot(tt_old, byy_old ,pch=20, xlab = "Time of Death (TOD)", ylab="Expression Value", main = title_old, ylim = c(ymin, ymax))
lines(tt_base, byy_old_fit, col = 4, lwd =3)
```


### Differential basal level

**Perform differential basal level analysis**
```{r}
result_diff_basal <- data.frame(genes = genes_both,
		basal_young = NA,
		basal_old = NA, 
		basal_diff = NA, 
		pvalue = NA
		)

for(gg in 1:nrow(result_diff_basal)){
  ayy_young <- BA11_young23[gg, ]
  ayy_old <- BA11_old23[gg, ]

  adiff_basal <- LR_diff(tt_young,ayy_young,tt_old,ayy_old,type="basal") # 

  result_diff_basal$basal_young[gg] <- adiff_basal$offset_1
  result_diff_basal$basal_old[gg] <- adiff_basal$offset_2
  result_diff_basal$basal_diff[gg] <- adiff_basal$offset_2 - adiff_basal$offset_1
  result_diff_basal$pvalue[gg] <- adiff_basal$pvalue
}

result_diff_basal_sorted <- result_diff_basal[order(result_diff_basal$pvalue), ]
```

**Save the result**

```{r, eval = FALSE}
write.csv(result_diff_basal_sorted, file = "result_diff_basal_sorted.csv", row.names = FALSE)
```

**Visualize the most significant gene**

```{r fig3, fig.height = 4, fig.width = 10}
topGene_diffbasal <- result_diff_basal_sorted$genes[1]
byy_young <- BA11_young23[topGene_diffbasal, ]
aLR_young <- LR_rhythmicity(tt_young, byy_young)
byy_young_fit <- sinCurve(tt_base, aLR_young$amp, aLR_young$phase, aLR_young$offset)
	
byy_old <- BA11_old23[topGene_diffbasal, ]
aLR_old <- LR_rhythmicity(tt_old, byy_old)
byy_old_fit <- sinCurve(tt_base, aLR_old$amp, aLR_old$phase, aLR_old$offset)

ymax <- max(c(byy_young, byy_old))
ymin <- min(c(byy_young, byy_old))

title_young <- paste0("young group, basal = ", signif(result_diff_basal_sorted$basal_young[1],2))
title_old <- paste0("old group, basal = ", signif(result_diff_basal_sorted$basal_old[1],2))

par(mfrow = c(1,2))
plot(tt_young, byy_young ,pch=20, xlab = "Time of Death (TOD)", ylab="Expression Value", main = title_young, ylim = c(ymin, ymax))
lines(tt_base, byy_young_fit, col = 2, lwd =3)

plot(tt_old, byy_old ,pch=20, xlab = "Time of Death (TOD)", ylab="Expression Value", main = title_old, ylim = c(ymin, ymax))
lines(tt_base, byy_old_fit, col = 4, lwd =3)
```

## Differential fit

### Candidate gene selection
The prerequisite for differential fit is  that  there  should  exist circadian  rhythmicity  in  either  young and old groups under comparisons.
Therefore, we suggested users to set $p \le 0.05$ from our previous likelihood-based test to ensure the existence of the circadian rhythmicity in either conditions.


```{r}
## genes that have circadian rhythmicity in either young and old groups
logic_either <- thisResult_old$pvalue<0.05 | thisResult_young$pvalue<0.05
genes_either <- thisResult_old$genes[logic_either]

BA11_old581 <- BA11_old[genes_either,]
BA11_young581 <- BA11_young[genes_either,]
```

### Differential fit

**Perform differential fit analysis**
```{r}
result_diff_fit <- data.frame(genes = genes_either,
		sigma2_young = NA,
		sigma2_old = NA, 
		sigma2_diff = NA, 
		pvalue = NA
		)

for(gg in 1:nrow(result_diff_fit)){
  ayy_young <- BA11_young581[gg, ]
  ayy_old <- BA11_old581[gg, ]

  adiff_fit <- LR_diff(tt_young,ayy_young,tt_old,ayy_old,type="fit") # 

  result_diff_fit$sigma2_young[gg] <- adiff_fit$sigma2_1
  result_diff_fit$sigma2_old[gg] <- adiff_fit$sigma2_2
  result_diff_fit$sigma2_diff[gg] <- adiff_fit$sigma2_2 - adiff_fit$sigma2_1
  result_diff_fit$pvalue[gg] <- adiff_fit$pvalue
}

result_diff_fit_sorted <- result_diff_fit[order(result_diff_fit$pvalue), ]
```

**Save the result**

```{r, eval = FALSE}
write.csv(result_diff_fit_sorted, file = "result_diff_fit_sorted.csv", row.names = FALSE)
```

**Visualize the most significant gene**

```{r fig4, fig.height = 4, fig.width = 10}
topGene_difffit <- result_diff_fit_sorted$genes[3]
byy_young <- BA11_young581[topGene_difffit, ]
aLR_young <- LR_rhythmicity(tt_young, byy_young)
byy_young_fit <- sinCurve(tt_base, aLR_young$amp, aLR_young$phase, aLR_young$offset)
	
byy_old <- BA11_old581[topGene_difffit, ]
aLR_old <- LR_rhythmicity(tt_old, byy_old)
byy_old_fit <- sinCurve(tt_base, aLR_old$amp, aLR_old$phase, aLR_old$offset)

ymax <- max(c(byy_young, byy_old))
ymin <- min(c(byy_young, byy_old))

title_young <- paste0("young group, sigma2 = ", signif(result_diff_fit_sorted$sigma2_young[1],2))
title_old <- paste0("old group, sigma2 = ", signif(result_diff_fit_sorted$sigma2_old[1],2))

par(mfrow = c(1,2))
plot(tt_young, byy_young ,pch=20, xlab = "Time of Death (TOD)", ylab="Expression Value", main = title_young, ylim = c(ymin, ymax))
lines(tt_base, byy_young_fit, col = 2, lwd =3)

plot(tt_old, byy_old ,pch=20, xlab = "Time of Death (TOD)", ylab="Expression Value", main = title_old, ylim = c(ymin, ymax))
lines(tt_base, byy_old_fit, col = 4, lwd =3)
```


```{r}
sessionInfo()
```
