---
 title: "RepeatedCircadian tutorial"
 auther: "Zhiguang Huo, Haocheng Ding (Department of Biostatistics, University of Florida)"
 date: "`r Sys.Date()`"
 output: 
   html_document:
    toc: true
    toc_depth: 2
    number_sections: true
 vignette: >
  %\VignetteIndexEntry{RepeatedCircadian}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(stringsAsFactors = FALSE)
```

# Prepare Example Data
```{r}
library(RepeatedCircadian) # include RepeatedCircadian package
# load the data
data("subHTRF")
data("pheno")
```

```{r}
Symbol <- subHTRF[,2]
ensemble <- row.names(subHTRF)
subHTRF <- subHTRF[,-c(1:3)]
ngene <- nrow(subHTRF)
sample <- colnames(subHTRF)
sample_split <- strsplit(sample, split=".", fixed=T)
ID <- unlist(lapply(sample_split, function(x) x[1]))
condition <- unlist(lapply(sample_split, function(x) x[2]))
time <- unlist(lapply(sample_split, function(x) x[3]))
sample_df <- data.frame(ID, condition, time)
sample_df$Time <- 1
sample_df$Time[which(sample_df$time=="T5")] <- 5
sample_df$Time[which(sample_df$time=="T9")] <- 9
sample_df$Time[which(sample_df$time=="T13")] <- 13
sample_df$Time[which(sample_df$time=="T17")] <- 17
sample_df$Time[which(sample_df$time=="T19")] <- 19

indR <- which(sample_df$condition=="R")  # 63
indU <- which(sample_df$condition=="U")  # 62




# Real Data circadian pattern detection 
ttR <- sample_df$Time[indR] # Circadian time of restricted group
ttU <- sample_df$Time[indU] # Circadian time of unrestricted group
idR <- sample_df$ID[indR] # ID of restricted group
idU <- sample_df$ID[indU] # ID of unrestricted group
```

# Circadian Rhythmicity Detection

## Perform circadian rhythmicity detection for human restricted feeding subdata

```{r}
# Circadian Analysis for Restricted Group
tableR <- matrix(nrow=ngene, ncol=7)
for (i in 1:ngene){
  yy <- as.numeric(subHTRF[i, indR])
  aTest <- rpt_rhythmicity(ttR, yy, idR, period=24, method="LR")
  tableR[i,] <- unlist(aTest)
}
tableR <- data.frame(ensemble,Symbol, tableR)
colnames(tableR)<-c("ensemble","gene name","statistic","pvalue","amplitude","phi","basal","sigma_alpha","sigma_0")
p.value <- tableR$pvalue
q.value <- p.adjust(p.value, method = "BH", n = length(p.value))
tableR <- data.frame(tableR, qvalue = q.value)
colnames(tableR)<-c("ensemble","geneName","statistic","pvalue","amplitude","phi","basal","sigma_alpha","sigma_0","qvalue")
tableR$r <- tableR$amplitude/tableR$sigma_0 # (sigmaA2: residual variance of the circadian model)


# Circadian Analysis for Unrestricted Group
tableU <- matrix(nrow=ngene, ncol=7)
for (i in 1:ngene){
  yy <- as.numeric(subHTRF[i, indU])
  aTest <- rpt_rhythmicity(ttU, yy, idU, period=24, method="LR")
  tableU[i,] <- unlist(aTest)
}

tableU <- data.frame(ensemble,Symbol,tableU)

colnames(tableU)<-c("ensemble","geneName","statistic","pvalue","amplitude","phi","basal","sigma_alpha","sigma_0")
p.value <- tableU$pvalue
q.value <- p.adjust(p.value, method = "BH", n = length(p.value))
tableU <- data.frame(tableU, qvalue = q.value)
colnames(tableU)<-c("ensemble","geneName","statistic","pvalue","amplitude","phi","basal","sigma_alpha","sigma_0","qvalue")
tableU$r <- tableU$amplitude/tableU$sigma_0 # (sigmaA2: residual variance of the circadian model)
```

## Save the Circadian Rhythmicity result as csv files.
```{r, eval = FALSE}
write.csv(tableR, "Table_pattern_Restricted_notorder.csv", row.names = F)
write.csv(tableU, "Table_pattern_Unrestricted_notorder.csv", row.names = F)
oR <- order(tableR$pvalue)
otableR <- tableR[oR,]
oU <- order(tableU$pvalue)
otableU <- tableU[oU,]
write.csv(otableR, "Table_pattern_Restricted.csv", row.names = F)
write.csv(otableU, "Table_pattern_Unrestricted.csv", row.names = F)
```


## Plot top significant genes with circadian rhythmicity in restricted group
```{r}
sinCurve <- function(x, A, phase, basal){
	A24 <- 24
	A * sin(2*pi/A24 * (x + phase)) + basal
}
tt_base <- seq(-6,24,0.1)

# Restricted group top 4 significant genes plots
topN <- 4
par(mfrow = c(2, 2))
for(i in 1:topN){
	this_LR <- otableR[i,]
	yy_obs <- subHTRF[this_LR$ensemble, indR]
	rptout <- rpt_rhythmicity(ttR,as.numeric(subHTRF[this_LR$ensemble, indR]),idR)
	yy_fit <- sinCurve(tt_base, rptout$A, rptout$phi, rptout$basal)
	atitle <- paste0(otableR[i,2], ", p = ", signif(this_LR$pvalue,3))	
	plot(ttR, yy_obs ,pch=20, xlab = "Circadian Time", ylab="Expression Value", main = atitle)
	lines(tt_base, yy_fit, col = 2, lwd =3)
}
```


## Plot top significant genes with circadian rhythmicity in unrestricted group
```{r}
sinCurve <- function(x, A, phase, basal){
	A24 <- 24
	A * sin(2*pi/A24 * (x + phase)) + basal
}
tt_base <- seq(-6,24,0.1)

# Unrestricted group top 4 significant genes plots
topN <- 4
par(mfrow = c(2, 2))
for(i in 1:topN){
	this_LR <- otableU[i,]
	yy_obs <- subHTRF[this_LR$ensemble, indU]
	rptout <- rpt_rhythmicity(ttU,as.numeric(subHTRF[this_LR$ensemble, indU]),idU)
	yy_fit <- sinCurve(tt_base, rptout$A, rptout$phi, rptout$basal)
	atitle <- paste0(otableU[i,2], ", p = ", signif(this_LR$pvalue,3))	
	plot(ttU, yy_obs ,pch=20, xlab = "Circadian Time", ylab="Expression Value", main = atitle)
	lines(tt_base, yy_fit, col = 2, lwd =3)
}
```

# Differential Circadian Pattern Detection

## Perform circadian rhythmicity detection for human restricted feeding subdata

```{r}
group <- c(rep("group1", length(idR)), rep("group2", length(idU)))

diff_table <- matrix(nrow=ngene, ncol=10)
for (i in 1:ngene){
  print(i)
  yyR <- as.numeric(subHTRF[i, indR])
  yyU <- as.numeric(subHTRF[i, indU])
  diff_out <- rpt_diff(ttR,yyR,ttU,yyU,idR,idU,group)
  diff_table[i,] <- unlist(diff_out)
}
diff_table_sum <- data.frame(ensemble,Symbol,diff_table)

colnames(diff_table_sum)<-c("ensemble","geneName","statistic","pvalue","amplitude1","amplitude2","phi1","phi2","basal1","basal2","sigma_alpha","sigma_0")
p.value <- diff_table_sum$pvalue
q.value <- p.adjust(p.value, method = "BH", n = length(p.value))
diff_table_sum <- data.frame(diff_table_sum, qvalue = q.value)
colnames(diff_table_sum)<-c("ensemble","geneName","statistic","pvalue","amplitude1","amplitude2","phi1","phi2","basal1","basal2","sigma_alpha","sigma_0","qvalue")
diff_table_sum$r1 <- diff_table_sum$amplitude1/diff_table_sum$sigma_0 # (sigmaA2: residual variance of the circadian model)
diff_table_sum$r2 <- diff_table_sum$amplitude2/diff_table_sum$sigma_0
```


## Plot top significant genes with differential circadian pattern
```{r}
sinCurve <- function(x, A, phase, basal){
	A24 <- 24
	A * sin(2*pi/A24 * (x + phase)) + basal
}
tt_base <- seq(-6,24,0.1)

# Unrestricted group top 4 significant genes plots
topN <- 4
par(mfrow = c(2, 2))
for(i in 1:topN){
	this_LR <- otableU[i,]
	yy_obs <- subHTRF[this_LR$ensemble, indU]
	rptout <- rpt_rhythmicity(ttU,as.numeric(subHTRF[this_LR$ensemble, indU]),idU)
	yy_fit <- sinCurve(tt_base, rptout$A, rptout$phi, rptout$basal)
	atitle <- paste0(otableU[i,2], ", p = ", signif(this_LR$pvalue,3))	
	plot(ttU, yy_obs ,pch=20, xlab = "Circadian Time", ylab="Expression Value", main = atitle)
	lines(tt_base, yy_fit, col = 2, lwd =3)
}
```

```{r}
sessionInfo()
```
